---
title: "In-class_Exercise 2"
execute: 
  warning: false
  message: false
editor: visual
---

## Overview

In this exercise, we are to find out if functionally similar water points positively co-occur in Nigeria.

## Getting Started

In the code chunk below, *p_load()* of **pacman** package is used to install and load the following R packages into R environment:

-   **sf**

-   **tidyverse**

-   **tmap**

-   **spdep**

-   **funModeling**, to be used for rapid Exploratory Data Analysis

```{r}
pacman::p_load(sf, tidyverse, tmap, spdep, funModeling)
```

## Import Data

change file name

-   geo_export (from the 4 shapefile data)

-   nga_admbnda_adm2_osgof_20190417

### Import water point geospatial data

The code chunk below uses...

```{r}
# eval: false
wp <- st_read(dsn = "rawdata",
              layer = "geo_export",
              crs = 4326) %>%
  filter(clean_coun == "Nigeria")
```

The code chunk below uses *write_rds()* of **readr** package...

```{r}
# eval: false
wp_nga <- write_rds(wp,
                    "rawdata/wp_nga.rds")
```

wpt_functional is a simple feature data table since wp_nga is a simple feature data table

remember to create the proportion! just divide the total will do.
